Parsing Postfix Log Files

### 2009-04-24 doc/logparser-other-parsers-review.tex
Changes made after proofreading the preface and chapters 1-3.

At the start of this project ten Postfix log file parsers were tested, with
the hope of finding a suitable parser to build upon, rather than writing
one from scratch.  It was quite difficult to find ten parsers to review for
this project, and the functionality offered by those parsers ranges from
quite basic to much more mature, depending on the needs of the author of
the parser.

It was hoped to reuse an existing parser rather than writing one from
scratch, but the existing parsers considered were rejected for one or more
reasons.  The effort required to adapt and improve an existing parser was
judged to be greater than the effort to write a new one, because the
techniques used by the existing parsers severely limited their potential:
some ignored the majority of log lines, parsing specific log lines
accurately, but without any provision for parsing new or similar log lines;
others sloppily parsed the majority of log lines, but were incapable of
distinguishing between log lines within one category, e.g.\ not
distinguishing between different anti-spam techniques causing delivery
attempts to be rejected.  The first parser
reviewed~\cite{log-mail-analyser} is the only previously published research
in this area that the author is aware of; that research aims to show that
providing the data from log files in a more accessible form is helpful to
systems administrators.

The ten parsers have been reviewed and compared with \parsername{}, this
project's parser, to show how much effort would have been required to use
one of those parsers to fulfil the aims and requirements of this project.
It is important to compare and contrast newly developed algorithms and
parsers against those already available, to accurately judge what
improvements, if any, are delivered by the newcomers.  \Tableref{Summary of
parsers' features} summarises the results of this review.

@@

    \item Only \parsername{} enables parsing of new log lines without
        extensive and intrusive modifications to the parser; the
        architecture that enables this is documented in \sectionref{parser
        architecture}.

    \item The reviewed parsers all produce a report of varying complexity
        and detail, whereas \parsername{} does not; it extracts data and
        leaves generation of reports to other programs.  Using a
        \acronym{SQL} database simplifies the process of generating such
        reports (discussed in \sectionref{database as API}); some sample
        queries are given in \sectionref{motivation}.  The parser developed
        for this project is designed to enable much more detailed log file
        analysis by providing a stable platform for subsequent programs to
        build upon.

@@

The data extracted by \parsername{} is documented in
\sectionref{connections table} and \sectionref{results table}; for
convenience that list is repeated here: client and server IP address and
hostname, HELO hostname, queueid, start time, end time, \acronym{SMTP}
code, enhanced status code, sender, recipient, size, message-id, delay, and
delays.  Note: \parsername{} correlates log lines by both queueid and
\acronym{pid} (\sectionref{queueid vs pid}), and it stores each mail's
queueid, but it does not store each connection's \acronyms{pid}, because a
use has not been identified for that data.

@@

There appears to be only one prior published paper about parsing Postfix
log files: \textit{Log Mail Analyzer: Architecture and Practical
Utilizations\/}~\cite{log-mail-analyser}.  The aim of \acronym{LMA} is
quite different from \parsername{}: it attempts to present correlated data
from log files in a form suitable for a systems administrator to search
using the myriad of standard Unix text processing utilities already
available.  It produces a \acronym{CSV} file and either a MySQL or Berkeley
DB database.  Hardly any documentation is provided with \acronym{LMA}, but
some documentation is available in~\cite{log-mail-analyser}.  Studying the
source code is informative, though this author had difficulty as the
authors of \acronym{LMA} wrote in Italian.

\acronym{CSV} is a very simple format where each record is stored in a
single line, with fields separated by a comma or other punctuation symbol.
Problems with \acronym{CSV} files include the need to escape separators in
the data stored, providing multiple values for a field (e.g.\ multiple
recipients), and adding new fields.  \acronym{CSV} files do not have a
standard mechanism to document the fields or the separator, unlike
\acronym{SQL} where every database includes a schema naming the fields and
the type of data they store (e.g.\ integer, text, timestamp).  The
\acronym{CSV} record format used by \acronym{LMA} is not documented, but
the output file contains a comment in Italian giving the format:\newline{}
\texttt{\# Timestamp|Nome Client|IP Client|IP Server|From|To|Status|Size}
\newline{}\acronym{LMA} treats \acronym{CSV} lines starting with
\texttt{\#} as comments, but not all \acronym{CSV} parsers will.

Berkeley DB only supports storing simple \textbf{(key, value)} pairs,
unlike \acronym{SQL} databases that store arbitrary tuples.  In
\acronym{LMA}'s main table the key is an integer referred to by secondary
tables, and the value is a \acronym{CSV} line containing all of the data
for that row.  The secondary by-sender, by-recipient, by-date, and by-IP
tables use the sender/recipient/date/IP address as the key, and the value
is a \acronym{CSV} list of integers referring to the keys in the main
table.  This effectively re-implements \acronym{SQL} foreign keys, but
without the functionality offered by even the most basic of \acronym{SQL}
databases, e.g.\ joins, ordering, searches.  It also requires custom code
to search on a combination of attributes, though the authors of
\acronym{LMA} did provide some simple reports: IP-STORY, FROM-STORY,
DAILY-EMAIL, and DAILY-REJECT (all described later).  Berkeley DB appears
to be the least useful of the three output formats: it does not provide the
functionality of a basic \acronym{SQL} database, and unlike \acronym{CSV}
files it cannot be used with standard Unix text processing tools.

The schema used with the MySQL database is undocumented, but at least it is
possible to discover the schema with an existing \acronym{SQL} database,
unlike with Berkeley DB\@; all \acronym{SQL} databases embed the schema
into the database and provide commands for displaying it.  Berkeley DB does
not embed a schema in its files, because it does not need one, and
embedding one would not provide any benefit; it only provides \textbf{(key,
value)} pairs, so any additional structuring of the data is imposed by the
application, and it is the responsibility of the author of the application
to document this structure.  MySQL support was not tested because the
schema is not documented, and no existing database was available to extract
the schema from, so the required tables could not be created.

Whether a MySQL database or Berkeley DB table is chosen in addition to the
\acronym{CSV} output, \acronym{LMA} stores the following data: time and
date of the log line, client hostname and IP address, server IP address,
sender and recipient addresses, \acronym{SMTP} code, and size (for accepted
mails only).  Unlike \parsername{} it does not store the server hostname,
HELO hostname, queueid, start and end times, timestamps for each log line,
enhanced status code, delivery delays, or message id (for accepted mails
only).  Handling of multiple recipients, \acronym{SMTP} codes, or remote
servers\footnote{A single mail may be sent to multiple remote servers if it
was addressed to recipients in different domains, or Postfix needs to try
multiple servers for one or more recipients.} is not explained;
experimental observation shows that multiple records are added when a mail
had multiple recipients, but the records are not associated or linked in
any way, and presumably the same approach was taken when there were
multiple destination servers.

@@

\acronym{LMA} does not deal with any of the complications discussed in
\sectionref{complications}, except for correlating log lines by queueid;
not correlating log lines by \acronym{pid} means it cannot correlate most
rejected delivery attempts.  It does not differentiate between different
types of rejections, so it is not suitable for the purposes of this
project; the data about which restriction caused the rejection is
discarded, whereas the main goal of this project is to retain that
information to aid optimisation and evaluation of anti-spam techniques.
\acronym{LMA} fails to parse Postfix log files generated on Solaris hosts
because the fields automatically prepended to each log line differ from
those prepended on Linux hosts; log files from Solaris hosts (and possibly
other operating systems) thus require pre-processing before parsing by
\acronym{LMA}.  The \numberOFlogFILES{} pre-processed test log files were
parsed without complaint by \acronym{LMA}, although it produced 32 entries
in the output \acronym{CSV} file for every rejection in the input log
files; it also missed some 40\% of delivered mail.  Once these deficiencies
were discovered the author did not spend any more time checking the
results.

@@

    \item \acronym{LMA} produces 32 lines in its \acronym{CSV} file for
        every single line it should have produced; the command above
        suppresses duplicate sequential lines.  DAILY-REJECT produces the
        correct number of output lines, probably because it uses the
        Berkeley DB files as input rather than the \acronym{CSV} file.

@@

    \item Some records output by DAILY-REJECT are truncated: they are
        missing the last $|$ that separates fields and also the newline
        following it, so the line containing only a single space is
        concatenated with the record.

@@

Pflogsumm produces a report designed for troubleshooting rather than
in-depth analysis.  It does not support saving any of the data it extracts
from log files, and it does not extract any data that it does not require
to produce its report: the HELO hostname, queueid, start and end times,
timestamps for each log line, or message id.  Both the parsing and
reporting are difficult to extend because it is a specialised tool, unlike
the easily extensible design of \parsername{}.  It does not correlate log
lines by queueid or \acronym{pid}, and does not need to deal with the
complications encountered during this project (\sectionref{complications}).
Pflogsumm produces a useful report, and successfully parsed the
\numberOFlogFILES{} log files it was tested with.  The results it reported
were not verified in detail, but it did not report any errors, and has an
excellent reputation amongst Postfix users.  Pflogsumm has many options to
include or exclude certain sections of the report, all clearly documented;
by default its report includes the following:

@@

\parserblurb{%
    Sawmill is a Postfix log analyzer (it also support 818 other log
    formats).  It can process log files in Postfix format, and generate
    dynamic statistics from them, analyzing and reporting events.  Sawmill
    can parse Postfix logs, import them into a SQL database (or its own
    built-in database), aggregate them, and generate dynamically filtered
    reports, all through a web interface.  Sawmill can perform Postfix
    analysis on any platform, including Window~[sic], Linux, FreeBSD,
    OpenBSD, Mac OS, Solaris, other UNIX, and more.
}
{http://www.thesawmill.co.uk/formats/postfix.html}
{2008/11/23}

@@

    \item
        \urlLastChecked[\hfill{}\newline{}]{http://www.thesawmill.co.uk/formats/postfix.html}{2008/11/23}.
        Fields extracted: from, to, server, UID, relay, status, number of
        recipients, origin hostname, origin IP address, and virus.  It also
        counts the number of and total size of all mails delivered.  The
        fields \texttt{server}, \texttt{uid}, \texttt{relay}, and
        \texttt{virus} are not explained in the documentation:
        \texttt{server} is probably the hostname or IP address of the
        server the mail is delivered to; \texttt{relay} might be the
        delivery method: \acronym{SMTP}, local delivery, or \acronym{LMTP};
        \texttt{uid} might be the uid of the user submitting mail locally.
        Postfix does not do any form of virus checking (though it has many
        options for cooperating with an external virus scanner), so the
        \texttt{virus} field is a mystery.

@@

    \item
        \urlLastChecked[\hfill{}\newline{}]{http://www.thesawmill.co.uk/formats/beta_postfix.html}{2008/11/23}.
        Fields extracted: from, to, client hostname, client IP address,
        relay hostname, relay IP address, status, response code, RBL list,
        and message id.  It also counts the number and size of all mails
        delivered, processed, blocked, expired, and bounced.

@@

When tested with the \numberOFlogFILES{} test log files it performed
adequately, though the rate it processed log files at did slow down
noticeably as it progressed.  Sawmill supports reading compressed log files
but it exhibits a dramatic slow down when doing so: it took six hours to
parse the first half of the log files, and twelve hours to parse the next
third; after twenty four hours spent parsing the remaining sixth it crashed
due to lack of disk space.  On the second parsing attempt the log files
were uncompressed beforehand and parsing took eight hours.

@@

The web interface is quite attractive and simple to use when searching, but
as an administrator it seems unnecessarily difficult to perform simple
tasks.  When testing Splunk it took roughly 30 minutes to figure out how to
add a single log file to be indexed for later searching, with the added
downside that the log file was copied into a spool directory before
indexing, doubling the disk space usage.  The next test was to index all
the log files in a particular directory, but after three hours, reading all
the available documentation, and numerous futile attempts, the author was
still unable to index all the log files in a directory using the web
interface.  Using Splunk's \acronym{CLI} rather than the web interface was
more successful: the command ``\texttt{splunk find logs
}\textit{log-directory\/}'' added 40 of the \numberOFlogFILES{} log files
to the queue for indexing.  Further attempts enqueued the same 40 log
files, without explaining why the others were excluded.\footnote{The log
files that were added multiple times appear to have been indexed once only;
presumably Splunk keeps track of the log files it has indexed and discards
requests to index log files for a second time.  This may or may not be a
useful feature for \parsername{}.}  The command did not have an option to
ensure the log files would be processed in the order they were created,
though such an option may be neither necessary nor beneficial with Splunk.
Subsequently the author was successful in adding a single log file at a
time using the command ``\texttt{splunk add tail }\textit{filename\/}''.  A
simple loop using that command was enough to add all the desired log files.
Splunk will periodically check all indexed log files for updates unless
they are manually removed from its list; this may or may not be useful
behaviour.  Splunk did not appear to have any difficulty in indexing the
log files once they had been successfully added to its queue.
\parsername{} parses the log files it is instructed to parse, in the order
they are given; periodic parsing of log files is a task an administrator
can easily achieve with \texttt{cron(8)} and \texttt{logrotate(8)}.

@@

Splunk supports reading compressed log files without any extra
configuration by the user, like \parsername{}.  The free version of Splunk
limits the volume of data indexed per day to 500MB, though a trial
Enterprise licence is available that allows indexing of up to 5GB of data
per day.  In 2007, the cheapest licenced version cost \$5000 plus \$1000
support per annum, and limited the volume of data indexed per day to
500MB\@.  Prices were removed from the Splunk website during 2008; now
Splunk's sales team must be asked for a quote.  Median log file size for
the log files used when evaluating \parsername{} is
\input{build/include-median-file-size}.

When parsing Postfix log files Splunk parses the standard
syslog\glsadd{syslog} fields at the beginning of the log line, and extracts
any \texttt{key=value} pairs occurring after the syslog prologue: to and
from addresses, HELO hostname, and protocol (\acronym{SMTP},
\acronym{LMTP}, or \acronym{ESMTP}).  \parsername{} extracts noticeably
more data (client and server IP address and hostname, queueid, start and
end times, timestamps for each log line, \acronym{SMTP} and enhanced status
codes, delivery delays, and message-id), though it does not make the full
text of the log line available; a few minutes work could add this to
\parsername{} if desired, but it would greatly increase the size of the
database.

@@

AWStats was not a suitable base for this project because it assumes that
all log files can be rewritten to be compatible with web server log files,
and will contain similar data; coercing Postfix log files into web server
log files, without substantial data loss, would require fully parsing the
Postfix log files without using AWStats, i.e.\ would require writing a
separate parser anyway.  It may be possible to use AWStats' graphing
capabilities to generate reports, by generating input for AWStats from the
database populated by \parsername{}.

@@

YAALA's Postfix log file parser extracts some of the fields \parsername{}
does: for client and server it stores either the IP address or the
hostname, not both; it omits the HELO hostname, queueid, \acronym{SMTP} and
enhanced status codes, size of each accepted mail, start and end times,
timestamps for each log line, and message-id\@.  It extracts one piece of
data that \parsername{} does not: how many times delivery was deferred for
each mail; this information can be determined from the database populated
by \parsername{} if desired.  Unlike \parsername{}, YAALA does not maintain
separate counters for different delivery attempt rejections, precluding the
possibility of using the collected data for optimisation, testing, or
understanding of restrictions.  YAALA's Postfix log file parser does not
deal with the complications explained in \sectionref{complications}, except
that it does correlate log lines by queueid.

@@

Testing YAALA was unsuccessful because all the select clauses tried
produced a similar error message:
\newline{}\tab{}\texttt{lib/Yaala/Data/Core.pm: Unavailable aggregation
requested:} \newline{}\tab{}\tab{}\texttt{``bytes''.  Returning 0.}
\newline{}  The underlying reason for this is that YAALA only parsed 408
(0.11\%) of 360,632 log lines in the first log file; it was not tested with
the remainder of the \numberOFlogFILES{} log files.

@@

Examining the source code reveals that Lire looks for
\texttt{<key>=<value>} pairs in each log line, extracts them, and
correlates the data by queueid.  This approach will extract the following
data: HELO hostname, queueid, \acronym{SMTP} code, sender and recipient
addresses, and size of accepted mails.  Lire misses the following fields
extracted by \parsername{}: client and server IP address and hostname,
start and end times, enhanced status code, delivery delays, timestamps of
each log line, and message-id\@.

Lire supports multiple output formats for the reports it generates (text,
HTML, PDF, and Excel 95) but the reports do not appear to be customisable,
and are not as detailed as Pflogsumm's; \parsername{} does not produce any
reports.  Lire supports saving extracted data for later report generation,
but the format of this data store is undocumented; given the source code it
should be possible, with enough time and effort, to understand the format.
\parsername{} uses an \acronym{SQL} database to make accessing the
extracted data as effortless as possible.  In general, Lire has poor
documentation.

Similar to AWStats and Logrep, Lire attempts to correlate log lines by
queueid, but not by \acronym{pid}, so the complete list of recipients for a
mail should be available; its parser extracts only part of the available
data and does not attempt to deal with the other complications described in
\sectionref{complications}.  When testing Lire on the \numberOFlogFILES{}
test log files it performed reasonably well: the numbers it reports appear
accurate, and the subset verified by the author were correct.  Its report
provides summaries of:

@@

Lire would not be a suitable base for this project: it does not extract
enough data; does not deal with rejected delivery attempts in any way; does
not make the extracted data easily available to other programs.  Its parser
is in-extensible; it could easily be replaced, but that would require
writing a parser from scratch, so would not be any less work.
\parsername{} could possibly be used to parse Postfix log files for Lire,
but the difficulty may outweigh the benefits, e.g.\ Lire's data store may
not be capable of storing the data extracted by \parsername{}, but the lack
of documentation hinders any evaluation.  As with YAALA this approach has
not been attempted by the author.

@@

        \item [\nichi{}] \parsername{} defers report generation to
            subsequent programs, but all the necessary data and
            documentation to produce reports is provided.

>>>>

Ten Postfix log file parsers were tested at the start of this project, with
the hope of finding a suitable parser to build upon, rather than writing
one from scratch.  It was quite difficult to find ten parsers to review,
and the functionality offered by those parsers ranges from quite basic to
more developed, depending on the needs of the parser's author.

It was hoped to reuse an existing parser rather than writing one from
scratch, but the existing parsers considered were rejected for one or more
reasons.  The effort required to adapt and improve an existing parser was
judged to be greater than the effort to write a new one, because the
techniques used by the existing parsers severely limited their potential:
some ignored the majority of log lines, parsing specific log lines
accurately, but without any provision for parsing new or similar log lines;
others sloppily parsed the majority of log lines, but were incapable of
distinguishing between log lines within one category, e.g.\ not
distinguishing between different anti-spam techniques causing delivery
attempts to be rejected.  The first parser
reviewed~\cite{log-mail-analyser} is the only previously published research
in this area that the author could find; that research aims to show that
providing the data from log files in a more accessible form is helpful to
systems administrators.

The ten parsers have been compared and contrasted with \parsername{}, this
project's parser, to show how much effort would have been required to use
one of those parsers to fulfil the aims and requirements of this project.
It is important to compare and contrast newly developed algorithms and
parsers against those already available, to accurately judge what
improvements, if any, are delivered by the newcomers.  \Tableref{Summary of
parsers' features} summarises the results of this review.

@@

    \item Only \parsername{} enables parsing of new log lines without
        extensive and intrusive modifications to the parser; the
        architecture enabling this is documented in \sectionref{parser
        architecture}.

    \item The reviewed parsers all produce a report of varying complexity
        and detail, whereas \parsername{} does not: it extracts data and
        leaves generation of reports to other programs.  Using a
        \acronym{SQL} database simplifies the process of generating such
        reports (discussed in \sectionref{database as API}); some sample
        queries are given in \sectionref{motivation}.  The parser developed
        for this project is designed to enable much more detailed log file
        analysis than other parsers by providing a stable platform for
        subsequent programs to build upon.

@@

The data extracted by \parsername{} is documented in
\sectionref{connections table} and \sectionref{results table}, but for
convenience the list is repeated here: client and server IP address and
hostname, HELO hostname, queueid, start time, end time, \acronym{SMTP}
code, enhanced status code, sender, recipient, size, message-id, delay, and
delays.  Note: \parsername{} correlates log lines by both queueid and
\acronym{pid} (\sectionref{queueid vs pid}), and it stores each mail's
queueid, but it does not store each connection's \acronym{pid}, because a
use has not been identified for that data.

@@

There appears to be only one prior published paper about parsing Postfix
log files: \textit{Log Mail Analyzer: Architecture and Practical
Utilizations\/}~\cite{log-mail-analyser}.  The aim of \acronym{LMA} is
quite different from \parsername{}: it attempts to present correlated data
from log files in a form suitable for a systems administrator to search
using the myriad of standard Unix text processing utilities already
available.  It produces a \acronym{CSV} file and either a MySQL or Berkeley
DB database.  Hardly any documentation is provided with \acronym{LMA}, but
some documentation is available in~\cite{log-mail-analyser}.  Studying the
source code is informative, though this author had some difficulty because
the authors of \acronym{LMA} wrote in Italian.

\begin{description}

    \item [CSV] \acronym{CSV} is a very simple format where each record is
        stored in a single line, with fields separated by a comma or other
        punctuation symbol.  Problems with \acronym{CSV} include the need
        to escape separators in the data stored, providing multiple values
        for a field (e.g.\ multiple recipients), and adding new fields.
        \acronym{CSV} does not have a standard mechanism to document the
        fields or the separator, unlike \acronym{SQL} where every database
        includes a schema naming the fields and the type of data they store
        (e.g.\ integer, text, timestamp).  The \acronym{CSV} record format
        used by \acronym{LMA} is not documented
        in~\cite{log-mail-analyser}, but the output file contains a comment
        in Italian giving the format:\newline{} \tab{} \texttt{\#
        Timestamp|Nome Client|IP Client|IP
        Server|From|To\newline{}\tab{}\tab{}|Status|Size}
        \newline{}\acronym{LMA} treats \acronym{CSV} lines starting with
        \texttt{\#} as comments, but not all \acronym{CSV} parsers will.

    \item [Berkeley DB] Berkeley DB only supports storing simple
        \textbf{(key, value)} pairs, unlike \acronym{SQL} databases that
        store arbitrary tuples.  In \acronym{LMA}'s main table the key is
        an integer referred to by secondary tables, and the value is a
        \acronym{CSV} line containing all of the data for that row.  The
        secondary by-sender, by-recipient, by-date, and by-IP tables use
        the sender/recipient/date/IP address as the key, and the value is a
        \acronym{CSV} list of integers referring to the keys in the main
        table.  This effectively re-implements \acronym{SQL} foreign keys,
        but without the functionality offered by even the most basic of
        \acronym{SQL} databases, e.g.\ joins, ordering, searches.  It also
        requires custom code to search on a combination of attributes, and
        the authors of \acronym{LMA} did provide some simple reports:
        IP-STORY, FROM-STORY, DAILY-EMAIL, and DAILY-REJECT (all described
        later).  Berkeley DB appears to be the least useful of the three
        output formats: it does not provide the functionality of a basic
        \acronym{SQL} database, and unlike \acronym{CSV} files it cannot be
        used with standard Unix text processing tools.

    \item [MySQL] The schema used with the MySQL database is undocumented,
        but at least it is possible to discover the schema if one has an
        existing \acronym{SQL} database, unlike with Berkeley DB\@; all
        \acronym{SQL} databases embed the schema into the database and
        provide commands for displaying it.  Berkeley DB does not embed a
        schema in its database files, because it does not need one, and
        embedding one would not provide any benefit; it only provides
        \textbf{(key, value)} pairs, so any additional structuring of the
        data is imposed by the application, and it is the responsibility of
        the author of the application to document this structure.
        \acronym['s]{LMA} MySQL support was not tested because the schema
        is not documented, and no existing database was available to
        extract the schema from, so the required database could not be
        created.

\end{description}

Whether a MySQL or Berkeley DB database is chosen in addition to the
\acronym{CSV} output, \acronym{LMA} stores the following data: time and
date of the log line, client hostname and IP address, server IP address,
sender and recipient addresses, \acronym{SMTP} code, and size (for accepted
mails only).  Unlike \parsername{} it does not store the server hostname,
HELO hostname, queueid, start and end times, timestamps for each log line,
enhanced status code, delivery delays, or message-id (for accepted mails
only).  Handling of multiple recipients, \acronym{SMTP} codes, or remote
servers\footnote{A single mail may be sent to multiple remote servers if it
was addressed to recipients in different domains, or Postfix needs to try
multiple servers for one or more recipients.} is not explained;
experimental observation shows that multiple records were added when a mail
had multiple recipients, but the records are not associated or linked in
any way, and presumably the same approach was taken when there were
multiple destination servers.

@@

\acronym{LMA} does not deal with any of the complications discussed in
\sectionref{complications}, except for correlating log lines by queueid; it
cannot correlate most rejected delivery attempts because it does not
correlate log lines by \acronym{pid}.  It does not differentiate between
different types of rejections, so it is not suitable for the purposes of
this project; the data about which restriction caused the rejection is
discarded, whereas the main goal of this project is to retain that
information to aid optimisation and evaluation of anti-spam techniques.
\acronym{LMA} fails to parse Postfix log files generated on Solaris hosts
because the fields automatically prepended to each log line differ from
those prepended on Linux hosts; log files from Solaris hosts (and possibly
other operating systems) thus require pre-processing before parsing by
\acronym{LMA}.  The \numberOFlogFILES{} pre-processed test log files were
parsed without complaint by \acronym{LMA}, although it produced 32 entries
in the output \acronym{CSV} file for every rejection in the input log
files; it also missed some 40\% of delivered mail.  Once these deficiencies
were discovered the author did not spend any more time checking the
results.

@@

    \item \acronym{LMA} produces 32 lines in its \acronym{CSV} file for
        every single line it should have produced; the command above
        suppresses duplicate sequential lines.  DAILY-REJECT produces the
        correct number of output lines, probably because it searches the
        Berkeley DB database rather than the \acronym{CSV} file.

@@

    \item Some records output by DAILY-REJECT are truncated: they are
        missing the last ``$|$'' that separates fields and also the newline
        following it, so the line containing only a single space is
        concatenated with the record.

@@

Pflogsumm produces a report designed for troubleshooting rather than
in-depth analysis.  It does not support saving any of the data it extracts
from log files, and it does not extract any data that it does not require
to produce its report: the HELO hostname, queueid, start and end times,
timestamps for each log line, or message-id.  Both the parsing and
reporting are difficult to extend because it is a specialised tool, unlike
the easily extensible design of \parsername{}.  It does not correlate log
lines by queueid or \acronym{pid}, and does not need to deal with the
complications encountered during this project (\sectionref{complications}).
Pflogsumm produces a useful report, and successfully parsed the
\numberOFlogFILES{} log files it was tested with.  The results it reported
were not verified in detail, but it did not report any errors, and has an
excellent reputation amongst Postfix users.  Pflogsumm has many options to
include or exclude certain sections of the report, all clearly documented;
by default its report includes the following:

@@

\parserblurb{%
    Sawmill is a Postfix log analyzer (it also support 818 other log
    formats).  It can process log files in Postfix format, and generate
    dynamic statistics from them, analyzing and reporting events.  Sawmill
    can parse Postfix logs, import them into a SQL database (or its own
    built-in database), aggregate them, and generate dynamically filtered
    reports, all through a web interface.  Sawmill can perform Postfix
    analysis on any platform, including Window, Linux, FreeBSD, OpenBSD,
    Mac OS, Solaris, other UNIX, and more.
}
{http://www.thesawmill.co.uk/formats/postfix.html}
{2008/11/23}

@@

    \item
        \urlLastChecked[\hfill{}\newline{}]{http://www.thesawmill.co.uk/formats/postfix.html}{2008/11/23}.
        Fields extracted: from, to, server, uid, relay, status, number of
        recipients, origin hostname, origin IP address, and virus.  It
        also counts the number of and total size of all mails delivered.
        The fields \texttt{server}, \texttt{uid}, \texttt{relay}, and
        \texttt{virus} are not explained in the documentation:
        \texttt{server} is probably the hostname or IP address of the
        server the mail is delivered to; \texttt{relay} might be the
        delivery method: \acronym{SMTP}, local delivery, or \acronym{LMTP};
        \texttt{uid} might be the uid of the user submitting mail locally.
        Postfix does not do any form of virus checking itself, so the
        \texttt{virus} field is a mystery.

@@

    \item
        \urlLastChecked[\hfill{}\newline{}]{http://www.thesawmill.co.uk/formats/beta_postfix.html}{2008/11/23}.
        Fields extracted: from, to, client hostname, client IP address,
        relay hostname, relay IP address, status, response code, RBL list,
        and message-id.  It also counts the number and size of all mails
        delivered, processed, blocked, expired, and bounced.

@@

When tested with the \numberOFlogFILES{} test log files it performed
adequately, though the rate it processed log files at did slowdown
noticeably as it progressed.  Sawmill supports reading compressed log files
but it exhibits a dramatic slow down when doing so: it took six hours to
parse the first half of the log files, and twelve hours to parse the next
third; after twenty four hours spent parsing the remaining sixth it crashed
due to lack of disk space.  On the second parsing attempt the log files
were uncompressed beforehand and parsing took eight hours.

@@

The web interface is quite attractive and simple to use when searching, but
as an administrator it seems unnecessarily difficult to perform simple
tasks.  When testing Splunk it took roughly 30 minutes to figure out how to
add a single log file to be indexed for later searching, with the added
downside that the log file was copied into a spool directory before
indexing, doubling the disk space usage.  The next test was to index all
the log files in a particular directory, but after three hours, reading all
the available documentation, and numerous futile attempts, the author was
still unable to index all the log files in a directory using the web
interface.  Using Splunk's command line interface rather than the web
interface was more successful: the command ``\texttt{splunk find logs
}\textit{log-directory\/}'' added 40 of the \numberOFlogFILES{} log files
to the queue for indexing.  Further attempts enqueued the same 40 log
files, without explaining why the others were excluded.\footnote{The log
files that were added multiple times appear to have been indexed once only;
presumably Splunk keeps track of the log files it has indexed and discards
requests to index log files for a second time.  This may or may not be a
useful feature for \parsername{}.}  The command did not have an option to
ensure the log files would be processed in the order they were created,
though such an option may be neither necessary nor beneficial with Splunk.
Subsequently the author was successful in adding a single log file at a
time using the command ``\texttt{splunk add tail }\textit{filename\/}'',
and then a simple loop using that command was enough to add all the desired
log files.  Splunk will periodically check all indexed log files for
updates unless they are manually removed from its list; this may or may not
be useful behaviour.  Splunk did not appear to have any difficulty in
indexing the log files once they had been successfully added to its queue.
\parsername{} parses the log files it is instructed to parse, in the order
they are given; periodic parsing of log files is a task an administrator
can easily achieve with \texttt{cron(8)} and \texttt{logrotate(8)}.

@@

Splunk supports reading compressed log files without any extra
configuration by the user, like \parsername{}.  The free version of Splunk
limits the volume of data indexed per day to 500MB, though a trial
Enterprise licence is available that allows indexing of up to 5GB of data
per day.  In 2007, the cheapest licenced version cost \$5000 plus \$1000
support per annum, and limited the volume of data indexed per day to
500MB\@.  Prices were removed from the Splunk website during 2008; now
Splunk's sales team must be asked for a quote.  Median log file size for
the \numberOFlogFILES{} log files used when evaluating \parsername{} is
\input{build/include-median-file-size}.

When parsing Postfix log files Splunk parses the standard
syslog\glsadd{syslog} fields at the beginning of the log line, and extracts
any \texttt{key=value} pairs occurring after the syslog prologue: to and
from addresses, HELO hostname, and protocol (\acronym{SMTP},
\acronym{LMTP}, or \acronym{ESMTP}).  \parsername{} extracts noticeably
more data: client and server IP address and hostname, queueid, start and
end times, timestamps for each log line, \acronym{SMTP} and enhanced status
codes, delivery delays, and message-id.  \parsername{} does not make the full
text of the log line available; a few minutes work could add this to
\parsername{} if desired, but it would greatly increase the size of the
database.

@@

AWStats was not a suitable base for this project because it assumes that
all log files can be rewritten to be compatible with web server log files,
and will contain similar data; coercing Postfix log files into web server
log files, without substantial data loss, would require fully parsing the
Postfix log files without using AWStats, i.e.\ would require writing a
separate parser anyway.  It may be possible to use AWStats' graphing
capabilities to generate reports, by generating input for AWStats from the
database populated by \parsername{}, but the author has not attempted that.

@@

YAALA's Postfix log file parser extracts some of the fields \parsername{}
does: for client and server it stores either the IP address or the
hostname, not both; it omits the HELO hostname, queueid, \acronym{SMTP} and
enhanced status codes, size of each accepted mail, start and end times,
timestamps for each log line, and message-id\@.  It extracts one piece of
data that \parsername{} does not: how many times delivery was deferred for
each mail; this information can be calculated from the database populated
by \parsername{} if desired.  Unlike \parsername{}, YAALA does not maintain
separate counters for different delivery attempt rejections, precluding the
possibility of using the collected data for optimisation, testing, or
understanding of restrictions.  YAALA's Postfix log file parser does not
deal with the complications explained in \sectionref{complications}, except
that it does correlate log lines by queueid (but not by \acronym{pid}).

@@

Testing YAALA was unsuccessful because all the queries produced a similar
error message:
\newline{}\tab{}\texttt{lib/Yaala/Data/Core.pm: Unavailable aggregation
requested:} \newline{}\tab{}\tab{}\texttt{``bytes''.  Returning 0.}
\newline{}  The underlying reason for this is that YAALA only parsed 408
(0.11\%) of 360,632 log lines in the first log file; it was not tested with
the remainder of the \numberOFlogFILES{} log files.

@@

Examining the source code reveals that Lire looks for
\texttt{<key>=<value>} pairs in each log line, extracts them, and
correlates the data by queueid (but not by \acronym{pid}).  This approach
will extract the following data: HELO hostname, queueid, \acronym{SMTP}
code, sender and recipient addresses, and size of accepted mails.  Lire
misses the following fields extracted by \parsername{}: client and server
IP address and hostname, start and end times, enhanced status code,
delivery delays, timestamps of each log line, and message-id\@.

Lire supports multiple output formats for the reports it generates (text,
HTML, PDF, and Excel 95) but the reports do not appear to be customisable,
and are not as detailed as Pflogsumm's; \parsername{} does not produce any
reports.  Lire supports saving extracted data for later report generation,
but the format of this data store is undocumented.  \parsername{} uses an
\acronym{SQL} database to make accessing the extracted data as effortless
as possible.  In general, Lire has poor documentation.

Similar to AWStats and Logrep, Lire attempts to correlate log lines by
queueid, but not by \acronym{pid}, so the complete list of recipients for
each delivered mail should be available; it does not attempt to deal with
the other complications described in \sectionref{complications}.  When
testing Lire on the \numberOFlogFILES{} test log files it performed
reasonably well: the numbers it reports appear accurate, and the subset
verified by the author were correct.  Its report provides summaries of:

@@

Lire would not be a suitable base for this project: it does not extract
enough data; does not deal with rejected delivery attempts in any way; does
not make the extracted data easily available to other programs.  Its parser
is not extensible; it could easily be replaced, but that would require
writing a parser from scratch, so would not be any less work than writing
\parsername{}.  \parsername{} could possibly be used to parse Postfix log
files for Lire, but the difficulty may outweigh the benefits, e.g.\ Lire's
data store may not be capable of storing the data extracted by
\parsername{}, but the lack of documentation hinders any evaluation.  As
with YAALA this approach has not been attempted by the author.

@@

        \item [\nichi{}] \parsername{} defers report generation to
            subsequent programs, but all the necessary documentation to
            produce reports is provided.
